Modeling HEPES buffer with a charge-balance approach

### 2023-11-17 manuscript.Rmd
Vectorize: partially describe the problem

### Vectorizing a function ???

JN: As yet I don't understand the workings of this, and must admit to never
having used Vectorize(). I think it worthwhile giving a good explanation.

>>>>

### Vectorizing a function

**IK**: This feels like I'm writing a tutorial. Should I be less
detailed?

**R** scripts naturally lend themselves to vectorization, thanks to the
language performing many operations on whole vectors. For example, **R**
has no problem evaluating the charge-balance equation function for a
vector of different values of $[\mathrm H]$ with other values remaining
scalars:

```{r HEPESFUNC}
HEPESFUNC <- function(H,SID,HEPTOT,pK1,pK2,pK3) {
	XX <- (H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
	IV <- HEPTOT/(1+XX)
	I <- IV*H^3/(10^-pK3*10^-pK2*10^-pK1)
	II <- IV*H^2/(10^-pK3*10^-pK2)
	III <- IV*H/10^-pK3
	H - kw/H + SID + I*2 + II - IV
}
```

This principle is used by the non-linear regression functions which, when
given a formula `y ~ f(x, theta)` and vectors `x` of independent
variables and `y` of observed values of the dependent variable, call the
model function `f` with the whole vector `x` as argument, expecting to
receive an estimate of the whole `y` at the current value of the
parameter `theta` in response.

Some functions are not naturally vectorized and cannot be easily changed
to obtain this property. For example, the model function for a
non-linear regression problem may involve root-finding using `uniroot`:

```{r}
# Which pH value satisfies the charge-balance equation?
pH <- function(SID, HEPTOT, pK1, pK2, pK3)
  -log10(uniroot(HEPESFUNC,c(1e-20,1),tol=1e-20,maxiter=1e4,
                 HEPTOT=HEPTOT,SID = SID, pK1=pK1,pK2=pK2,pK3=pK3)$root)
```

The underlying mechanism of `uniroot()` only works with scalar functions
of one argument, so passing vectors of length more than `1` to this
function will cause an error: `uniroot()` is not prepared to find
multiple different roots for every given set of the parameter values. In
order to use this function with non-linear regression methods, a wrapper
is needed. This wrapper must call the original function in a loop for
every element of the vector arguments passed to it:

```{r}
pH.wrapper <- function(SID, HEPTOT, pK1, pK2, pK3)
 vapply(SID, function(SID_i)
  -log10(uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
                 HEPTOT = HEPTOT, SID = SID_i,
                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root),
  numeric(1)
 )
```

The base function `Vectorize` automates this transformation. It returns
a wrapper function that obtains the values of every argument given to
the wrapper and passes them to a live-constructed call to `mapply()`,
letting the latter call the original function. A call to
`mapply(f, a, b, c)`, returns a vector of values of
`f(a[i], b[i], c[i])` for every applicable `i`, subject to vector
recycling rules.
